---
title: "Gov 50 Final Project"
author: "Tosca Langbert"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Final Project Milestone 3
 
For this final project, I am exploring a dataset containing information about billionaires worldwide. In particular, I am interested in exploring what factors are most strongly correlated with being a billionaire (e.g. whether the billionaire is self-made or conditions in where the billionaire lives) to better understand how these individuals have accrued so much wealth. 

```{r, echo=FALSE}
# Setup chunk
library(tidyverse)
library(dplyr)
library(ggplot2)
library(infer)
```

```{r, echo=FALSE}
billionaires <- read.csv("Billionaires Statistics Dataset.csv")
```

In the data visualization below, I'm comparing the net worths of self-made and non-self-made billionaires. Interestingly, there is a marked difference in their avaerage net worths--it looks like the typical self-made billionaire can expect to a net worth approximately $505 million (USD) lower than their counterparts with inherited wealth. Moving forward, I would like to explore whether there are confounding variables at play (for example, maybe people with inherited wealth are more likely to work in different industries). 

```{r, echo=FALSE}
ggplot(data = billionaires, 
       aes(x = selfMade, 
           y = finalWorth, 
           fill = selfMade)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar_format(prefix = "$")) +  # Change scale of y axis w/ help of GPT <3 
  labs(x = "Self-Made Status", 
       y = "Net Worth (in millions of dollars)", 
       title = "Comparing the Net Worth of Self-Made vs. Not Self-Made Billionaires", 
       fill = "Self-Made")
```

```{r, echo=FALSE}
billionaires |>
  group_by(selfMade) |>
  summarize(mean(finalWorth), sd(finalWorth)) |>
  knitr::kable(digits = 2,
               col.names = c("Self-Made Status", 
                             "Average Net Worth (in millions USD)",
                             "Net Worth Standard Dev."))
```


