---
title: "Gov 50 Final Project"
author: "Tosca Langbert"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---
In this project, I explore data about billionaires worldwide. 
I am interested in understanding what factors differentiate the "nouveau riche" and billionaires who inherited their wealth. Some variables I explore include whether there are specific industries or regions that are more strongly correlated with inherited on self-earned wealth.

```{r, echo=FALSE}
# Setup chunk
library(tidyverse)
library(dplyr)
library(ggplot2)
library(infer)
```
```{r, echo=FALSE}
billionaires <- read.csv("Billionaires Statistics Dataset.csv")
```
In the data visualization below, I compare the net worths of self-made and non-self-made billionaires. Interestingly, there is a marked difference in their avaerage net worths. The typical self-made billionaire can expect to have a net worth of approximately $505 million (USD) lower than their counterparts with inherited wealth.

```{r, echo=FALSE}
ggplot(data = billionaires, 
       aes(x = selfMade, 
           y = finalWorth, 
           fill = selfMade)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar_format(prefix = "$")) +  # Change scale of y axis w/ help of GPT <3 
  labs(x = "Self-Made Status", 
       y = "Net Worth (in millions of dollars)", 
       title = "Comparing the Net Worth of Self-Made vs. Not Self-Made Billionaires", 
       fill = "Self-Made") +
  theme_linedraw() +
  scale_fill_manual(values = c("tan1", "lightblue1"))
```

```{r, echo=FALSE}
billionaires |>
  group_by(selfMade) |>
  summarize(mean(finalWorth), sd(finalWorth)) |>
  knitr::kable(digits = 2,
               col.names = c("Self-Made Status", 
                             "Average Net Worth (in millions USD)",
                             "Net Worth Standard Dev."))
```

```{r, echo = FALSE}
ggplot(data = billionaires, 
       aes(x = gender, 
           y = finalWorth, 
           fill = gender)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar_format(prefix = "$")) +  # Change scale of y axis w/ help of GPT <3 
  labs(x = "Gender", 
       y = "Net Worth (in millions of dollars)", 
       title = "Comparing the Net Worth of Billionaires by Gender", 
       fill = "Gender") +
  theme_linedraw() +
  scale_fill_manual(values = c("pink1", "lightblue2"))
```
```{r, echo=FALSE}
billionaires |>
  group_by(gender) |>
  summarize(mean(finalWorth), sd(finalWorth)) |>
  knitr::kable(digits = 2,
               col.names = c("Gender", 
                             "Average Net Worth (in millions USD)",
                             "Net Worth Standard Dev."))
```

```{r, echo = FALSE}
# after install.packages("countrycode")
library(countrycode)

# Add continent variable
billionaires <- billionaires |>
  mutate(continent = countrycode(country, "country.name", "continent"))
```
```{r, echo = FALSE}
ggplot(data = billionaires, 
       aes(x = continent, 
           y = finalWorth, 
           fill = selfMade)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar_format(prefix = "$")) +  # Change scale of y axis w/ help of GPT <3 
  labs(x = "Self-Made Status", 
       y = "Net Worth (in millions of dollars)", 
       title = "Comparing Self-Made and Not Self-Made Billionaires by Continent", 
       fill = "Self-Made") +
  theme_linedraw() +
  scale_fill_manual(values = c("tan1", "lightblue1"))
```

```{r, echo = FALSE}
ggplot(data = billionaires, 
       aes(x = reorder(industries, -finalWorth, FUN = median),
           y = finalWorth, 
           fill = selfMade)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar_format(prefix = "$")) +
  labs(x = "Industry of Billionaire", 
       y = "Net Worth of Billionaire (in millions of dollars)", 
       title = "Comparing Billionaire Net Worth by Industry & Self-Made Status", 
       fill = "Self-Made Status") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle=45, hjust = 1)) +
  scale_fill_manual(values = c("tan1", "lightblue1"))
```

Source: https://www.kaggle.com/datasets/nelgiriyewithana/billionaires-statistics-dataset
