---
title: "Who becomes a billionaire and why?"
author: "Tosca Langbert"
description: "Gov 50 Final Project"
output:
  distill::distill_article:
    self_contained: false
---
## Introduction and Research Question

In this project, I explore data about billionaires worldwide. 
I am interested in understanding what factors differentiate the "nouveau riche" and billionaires who inherited their wealth. Some variables I explore include whether there are specific industries or regions that are more strongly correlated with inherited on self-earned wealth.

```{r, echo=FALSE}
# Setup chunk
library(tidyverse)
library(dplyr)
library(ggplot2)
library(infer)
library(knitr)
```

## Data (Source and Variables)

```{r, echo=FALSE}
billionaires <- read.csv("Billionaires Statistics Dataset.csv")
```

```{r, echo = FALSE}
# Distribution of net worth
ggplot(data = billionaires,
       mapping = aes(x = finalWorth)) +
  geom_histogram() +
  theme_minimal() +
  labs(x = "Net Worth (in millions USD)",
       y = "Count")
```
```{r}
summary(billionaires)
```


## Results

In the data visualization below, I compare the net worths of self-made and non-self-made billionaires. Interestingly, there is a marked difference in their avaerage net worths. The typical self-made billionaire can expect to have a net worth of approximately $505 million (USD) lower than their counterparts with inherited wealth.

```{r, echo=FALSE}
ggplot(data = billionaires, 
       aes(x = selfMade, 
           y = finalWorth, 
           fill = selfMade)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar_format(prefix = "$")) +  # Change scale of y axis w/ help of GPT <3 
  labs(x = "Self-Made Status", 
       y = "Net Worth (in millions of dollars)", 
       title = "Comparing the Net Worth of Self-Made vs. Not Self-Made Billionaires", 
       fill = "Self-Made") +
  theme_linedraw() +
  scale_fill_manual(values = c("tan1", "lightblue1"))
```

```{r, echo=FALSE}
billionaires |>
  group_by(selfMade) |>
  summarize(mean(finalWorth), median(finalWorth), sd(finalWorth)) |>
  knitr::kable(digits = 2,
               col.names = c("Self-Made Status", 
                             "Average Net Worth (in millions USD)",
                             "Median Net Worth (in millions USD)",
                             "Net Worth Standard Dev."))
```

```{r, echo = FALSE}
ggplot(data = billionaires, 
       aes(x = gender, 
           y = finalWorth, 
           fill = gender)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar_format(prefix = "$")) +  # Change scale of y axis w/ help of GPT <3 
  labs(x = "Gender", 
       y = "Net Worth (in millions of dollars)", 
       title = "Comparing the Net Worth of Billionaires by Gender", 
       fill = "Gender") +
  theme_linedraw() +
  scale_fill_manual(values = c("pink1", "lightblue3"))
```

```{r, echo=FALSE}
billionaires |>
  group_by(gender) |>
  summarize(mean(finalWorth), median(finalWorth), sd(finalWorth)) |>
  knitr::kable(digits = 2,
               col.names = c("Gender", 
                             "Average Net Worth (in millions USD)",
                             "Median Net Worth (in millions USD",
                             "Net Worth Standard Dev."))
```

```{r, echo = FALSE}
ggplot(data = billionaires, 
       aes(x = gender, 
           y = finalWorth, 
           fill = selfMade)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar_format(prefix = "$")) +  # Change scale of y axis w/ help of GPT <3 
  labs(x = "Gender", 
       y = "Net Worth (in millions of dollars)", 
       title = "Comparing the Net Worth of Billionaires by Gender and Self-Made Status", 
       fill = "Self-Made") +
  theme_linedraw() +
  scale_fill_manual(values = c("tan1", "lightblue1"))
```

```{r, echo = FALSE}
# after install.packages("countrycode")
library(countrycode)

# Add continent variable
billionaires <- billionaires |>
  mutate(continent = countrycode(country, "country.name", "continent"))
```
```{r, echo = FALSE}
ggplot(data = billionaires, 
       aes(x = continent, 
           y = finalWorth, 
           fill = selfMade)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar_format(prefix = "$")) +  # Change scale of y axis w/ help of GPT <3 
  labs(x = "Self-Made Status", 
       y = "Net Worth (in millions of dollars)", 
       title = "Comparing Self-Made and Not Self-Made Billionaires by Continent", 
       fill = "Self-Made") +
  theme_linedraw() +
  scale_fill_manual(values = c("tan1", "lightblue1"))
```

```{r, echo = FALSE}
ggplot(data = billionaires, 
       aes(x = reorder(industries, -finalWorth, FUN = median),
           y = finalWorth, 
           fill = selfMade)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar_format(prefix = "$")) +
  labs(x = "Industry of Billionaire", 
       y = "Net Worth of Billionaire (in millions of dollars)", 
       title = "Comparing Billionaire Net Worth by Industry & Self-Made Status", 
       fill = "Self-Made Status") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle=45, hjust = 1)) +
  scale_fill_manual(values = c("tan1", "lightblue1"))
```

```{r, echo = FALSE}
billionaires |>
  mutate(
    age_cohort = case_when(
      age < 20 ~ "Under 20",
      age >= 20 & age < 40 ~ "Between 20 and 39",
      age >= 40 & age < 60 ~ "Between 40 and 59",
      age >= 60 & age < 80 ~ "Between 60 and 79",
      age >= 80 ~ "Age 80 and Above",
    ), na.rm = TRUE) |>
  ggplot(mapping = aes(x = fct_reorder(age_cohort, age), 
                       fill = selfMade)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("tan1", "lightblue1")) +
  labs(title = "Frequency of Billionaires by Age Cohort and Self-Made Status",
       y = "Count",
       x = "Age Cohort",
       fill = "Self-Made Status")
## , y = fct_reorder(c("Under 20", "Between 20 and 39", "Between 40 and 59", "Between 60 and 79", "Age 80 and Above", "NA")
```


## Conclusion



## Bibliography

Source: https://www.kaggle.com/datasets/nelgiriyewithana/billionaires-statistics-dataset
